---
title: "Data_Science"
author: "Saurabh Gupta"
Date: 27 Aug 2025
format: html
editor: visual
---

## 

# Part 1 — Data Science

**What do you think being a data scientist is about?**\
A data scientist is about using data to **find patterns to support the decision making**. A data scientist collect and analyze the raw data to interpret information out of it. Often blending statistical programming to obtain the information and making some predictions.\
The work usually involves data prepare and clean up, looking for trends in it, and translating those findings into accurate and useful insights to support the decision making process.

**What differences/similarities do you see between data scientists and statisticians?**\
Both data scientists and statisticians work with data to answer questions and make sense of information.  They share the common task of **collecting data and analyzing it** to find insights.   Statisticians usually concentrate on applying **mathematical and statistical models**, creating studies, and making sure results are accurate and reliable.\
Data scientists, meanwhile, not only analyze but also manage and extract data, often building systems or tools that automate the process and turn findings into practical results.\
In simple terms, statisticians lean toward careful mathematical modeling and interpretation, while data scientists blend those methods with programming to create solutions that can be scaled and applied in real-world settings.

**How do you view yourself in relation to these two areas?**\
I am having a good understanding of both statistics and data science since I have studied them both in detail.\
From the **statistician’s side**, I apply methods like hypothesis testing and probability to make sound decisions.\
From the **data scientist’s side**, I like collecting ad leaning data, performing exploratory analysis, creating models, and generating predictions that lead to actionable insights.\
My long term goal is to bring these two areas together and apply them in the field of **Artificial Intelligence.**

# Part 2 — Data Science

### Add the iris data set

```{r}
# Add the iris dataset
y <- density(iris$Petal.Width)
```

### Find the class of y object

```{r}
# Class of y object
class(y)

```

### Find the type of y object

```{r}
# Type of y object 
typeof(y)
```

### Find the structure of y object

```{r}
# Structure of y object 
structure(y)

```

### Plot the graph on iris data set

```{r}
#| echo: false
plot(y, main = "Density of Iris Petal.Width")
```

# End of Data Science Homework 01

```         
```
