---
title: "seol_bike_modeling_practice"
editor: visual
---

```{r}


library(tidyverse)
library(tidymodels)
library(readr)
library(lubridate)
library(baguette)
library(rpart.plot)
library(rpart)
library(tree)
library(ranger)
library("vip")
library("parsnip")


```

```{r}
bike_data <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/SeoulBikeData.csv",local = locale(encoding = "latin1"))

bike_data
```

```{r}
 bike_data |> 
  is.na() |> 
  colSums()
```

There are no missing values in any column, so we don’t need to do any imputation.

```{r}
attributes(bike_data)$spec
```

```{r}
# Change date column 
bike_data <- bike_data |>
  mutate(date = dmy(Date)) |>
  select(-Date)
```

```{r}
summary(bike_data)
```

The bike counts range from 0 to about 3,556 rentals per hour, and temperatures range from about -18°C to 39°C

```{r}
print(bike_data$Seasons |> 
        unique())
```

```{r}
bike_data$Holiday |>
  unique()
```

```{r}
bike_data$`Functioning Day` |>
  unique()
```

The data spans all four seasons, includes both holidays and non-holidays, and some days where the system was not functioning.

### Turn it into factor to categorical data

```{r}
bike_data <- bike_data |>
  mutate(seasons = factor(Seasons),
         holiday = factor(Holiday),
         fn_day = factor(`Functioning Day`)) |>
  select(-Seasons, -Holiday, - `Functioning Day`)
```

```{r}
bike_data <- bike_data |> rename('bike_count' = `Rented Bike Count`,
         'hour' = "Hour",
         "temp" = `Temperature(°C)`,
         "wind_speed" = `Wind speed (m/s)`,
         "humidity" = `Humidity(%)`,
         "vis" = `Visibility (10m)`,
         "dew_point_temp" = `Dew point temperature(°C)`,
         "solar_radiation" = `Solar Radiation (MJ/m2)`,
         "rainfall" = "Rainfall(mm)",
         "snowfall" = `Snowfall (cm)`)
```

```{r}
bike_data <- bike_data |>
  filter(fn_day == "Yes") |>
  select(-fn_day)
```

```{r}
bike_data <- bike_data |>
  group_by(date, seasons, holiday) |>
  summarize(bike_count = sum(bike_count),
            temp = mean(temp),
            humidity = mean(humidity),
            wind_speed = mean(wind_speed),
            vis = mean(vis),
            dew_point_temp = mean(dew_point_temp),
            solar_radiation = mean(solar_radiation),
            rainfall = sum(rainfall),
            snowfall = sum(snowfall))|>
  ungroup()
```

Aggregated the data to the daily level and now have 353 days of total bike rentals with averaged weather and daily rainfall/snowfall totals.

```{r}
bike_data
```

### 

Summary Stats & Graphs

```{r}
bike_data |>
  summarize(across('bike_count',
                   .fns = c("mean "= mean,
                            "median" = median,
                            "sd" = sd,
                            "IQR" = IQR,
                            "min" = min,
                            "max" = max),
                   .names = "{.col}_{.fn}"))
```

On average there are about 17,500 bike rentals per day, with a median around 18,500 and a wide spread in daily counts.

```{r}
bike_data |>
  group_by(holiday)|>
  summarize(across('bike_count',
                   .fns = c("mean "= mean,
                            "median" = median,
                            "sd" = sd,
                            "IQR" = IQR,
                            "min" = min,
                            "max" = max),
                   .names = "{.col}_{.fn}"))
```

Daily rentals are lower on holidays than on non-holidays, and they also show more variability on holidays.

```{r}
bike_data |>
  group_by(seasons)|>
  summarize(across('bike_count',
                   .fns = c("mean "= mean,
                            "median" = median,
                            "sd" = sd,
                            "IQR" = IQR,
                            "min" = min,
                            "max" = max),
                   .names = "{.col}_{.fn}"))
```

```{r}
bike_data |>
  group_by(seasons,holiday) |>
  summarize(across('bike_count',
                   .fns = c("mean "= mean,
                            "median" = median,
                            "sd" = sd,
                            "IQR" = IQR,
                            "min" = min,
                            "max" = max),
                   .names = "{.col}_{.fn}"))
```

Summer and autumn have the highest daily bike use, while winter has much fewer rentals per day.

Within each season, non-holiday days usually have more rentals than holiday days, especially in spring and winter.

```{r}
bike_data |>
  select(where(is.numeric)) |>
  cor() |>
  round(3)
```

Bike rentals are strongly positively related to temperature, solar radiation, and dew point temperature, and negatively related to rainfall and snowfall. 

```{r}
ggplot(bike_data, aes(x = temp, y = bike_count)) +
  geom_jitter(aes(color = seasons)) +
  facet_grid(~holiday)
                
```

Warmer days tend to have more bike rentals, especially on non-holiday days.

```{r}
ggplot(bike_data, aes(x = solar_radiation, y = bike_count)) +
  geom_jitter(aes(color = seasons)) +
  facet_grid(~holiday)
```

Days with more sunlight (higher solar radiation) usually have higher bike usage.

```{r}
set.seed(23)
bike_split <- initial_split(bike_data, prop = 0.75, strata = seasons)
bike_train <- training(bike_split)
bike_test <- testing(bike_split)
bike_10_fold <- vfold_cv(bike_train, 10)
 
```

We split the data into 75% training and 25% test sets and created 10-fold cross-validation folds on the training data.

```{r}
MLR_rec1 <- recipe(bike_count ~., data = bike_train) |>
  step_date(date, features = "dow") |>
  step_mutate(day_type = factor(if_else(date_dow %in% c("Sat", "Sun"), "Weekend", "Weekday"))) |>
  step_rm(date, date_dow) |>
  step_dummy(seasons, holiday, day_type) |>
  step_normalize(all_numeric(), -bike_count)
```

```{r}
MLR_rec2 <- MLR_rec1 |>
  step_interact(terms = ~starts_with("seasons")*starts_with("holiday") +
                  starts_with("seasons")*temp +
                  temp*rainfall)
```

```{r}
MLR_rec3 <- MLR_rec2 |>
  step_poly(temp,
            wind_speed,
            vis,
            dew_point_temp,
            solar_radiation,
            rainfall,
            snowfall,
            degree = 2)

#MLR_rec3

```

```{r}
MLR_spec <- linear_reg() |>
  set_engine("lm")
```

```{r}
MLR_CV_fit1 <- workflow() |>
  add_recipe(MLR_rec1) |>
  add_model(MLR_spec) |>
  fit_resamples(bike_10_fold)
```

```{r}
MLR_CV_fit2 <- workflow() |>
  add_recipe(MLR_rec2) |>
  add_model(MLR_spec) |>
  fit_resamples(bike_10_fold)
```

```{r}
MLR_CV_fit3 <- workflow() |>
  add_recipe(MLR_rec3) |>
  add_model(MLR_spec) |>
  fit_resamples(bike_10_fold)
  
```

```{r}
rbind(MLR_CV_fit1 |> collect_metrics(),
      MLR_CV_fit2 |> collect_metrics(),
      MLR_CV_fit3 |> collect_metrics())
```

```{r}
rbind(MLR_CV_fit1 |> collect_metrics() |> filter(.metric == "rmse"),
      MLR_CV_fit2 |> collect_metrics() |> filter(.metric == "rmse"),
      MLR_CV_fit3 |> collect_metrics() |> filter(.metric == "rmse")) |>
  mutate(Model = c("Model1 ", "Model 2", "Model 3")) |>
  select(Model, mean, n, std_err)
```

All three linear models perform reasonably well, with R² values above 0.83 and RMSE between about 2,750 and 4,200 using cross-validation.

```{r}
final_fit <- workflow() |>
  add_recipe(MLR_rec1) |>
  add_model(MLR_spec) |>
  last_fit(bike_split)

```

```{r}
final_fit |>
  collect_metrics()
```

On the test set, the chosen linear model has an RMSE of about 3,850 and explains around 84% of the variation in daily bike rentals.

```{r}
final_fit |>
  extract_fit_parsnip() |>
  tidy()
```

The linear model suggests that higher solar radiation increases rentals, while more rainfall and winter season indicators decrease rentals, holding other variables constant.

### LASSO model

\*Create a Model instance with tune()

```{r}
LASSO_spec <- linear_reg(penalty = tune(), mixture = 1) |>
  set_engine("glmnet")
```

-   create a entire workflow

```{r}
LASSO_wkf <- workflow() |>
  add_recipe(MLR_rec1) |>
  add_model(LASSO_spec)

LASSO_wkf
  
```

\*n Fit the model with tune_grid and grid_regular

```{r}
LASSO_grid <- LASSO_wkf |>
  tune_grid(resamples = bike_10_fold,
            grid = grid_regular(penalty(), levels = 200))
LASSO_grid
```

-   

```{r}
LASSO_grid[1, ".metrics"][[1]]
```

```{r}
LASSO_grid |>
  collect_metrics() |>
  filter(.metric == "rmse")
```

The LASSO model performs similarly to the standard linear model but with slightly higher RMSE, so it does not clearly improve prediction.

-   Pull out the best Model

```{r}
lowest_rmse <- LASSO_grid |>
  select_best(metric = "rmse")

lowest_rmse
```

-   Fit the best LASSO model on training set

```{r}
LASSO_final <- LASSO_wkf |>
  finalize_workflow(lowest_rmse) |>
  fit(bike_train)

  
```

-   Coefficient from LASSO model

```{r}
 LASSO_final |>
   extract_fit_parsnip() |>
   tidy()

```

LASSO shrinks some coefficients toward zero (for example temperature), keeping only the most important effects while still capturing season, weather, and day-type patterns.

## Regression Tree Model

```{r}
tree_mod <- decision_tree(tree_depth = tune(),
                          min_n = tune(),
                          cost_complexity = tune()) |>
  set_engine("rpart") |>
  set_mode("regression")

```

-   Create a workflow

```{r}
tree_wkf <- workflow() |>
  add_recipe(MLR_rec1) |>
  add_model(tree_mod)

#tree_wkf
```

-   Tune regression tree

```{r}
tree_grid <- grid_regular(cost_complexity(),
                          tree_depth(range = c(2,10)),
                          min_n(range = c(5,20)),
                          levels = 5)
```

-   Perform cross-validation

```{r}
tree_tune <- tree_wkf |>
  tune_grid(
    resamples = bike_10_fold,
    grid = tree_grid,
    metrics = metric_set(rmse, mae)
  )

#tree_tune
```

```{r}
prep(MLR_rec1) |> bake(new_data = bike_train)
```

-   Select the best tree Model

```{r}
best_tree <- tree_tune |> 
  select_best(metric = "rmse")
#tree_best_params <- select_best(tree_tune, metric = "rmse")
best_tree

```

-   Fit best model on Training set

```{r}
final_tree_wkf <- tree_wkf |>
  finalize_workflow(best_tree)

# Fit final mode on the training set
tree_fit <- final_tree_wkf|>
  fit(data = bike_train)
```

Evaluate Model Performance on the Test Set

```{r}
# Predict on the test set
tree_test_results <- tree_fit %>%
  predict(new_data = bike_test) %>%
  bind_cols(bike_test) %>%
  metrics(truth = bike_count, estimate = .pred)

tree_test_results

```

The single regression tree achieves an RMSE of about 3,410 and R² around 0.87 on the test set, better than the linear models.

```{r}
```

-   Plot the tree

```{r}
library(rpart.plot)

tree_fit %>%
  extract_fit_engine() %>%      
  rpart.plot::rpart.plot(roundint = FALSE)


```

The regression tree splits the data into groups based mainly on weather and season to predict daily bike rentals.

## Tuned Bagged Tree Model

```{r}
bagged_tree <- bag_tree(tree_depth = tune(),
                        min_n = tune()) |>
  set_engine("rpart") %>%
  set_mode("regression")
```

# Fit a bag model

```{r}

```

-   Workflow

```{r}
bag_wkf <- workflow() |>
  add_recipe(MLR_rec1) |>
  add_model(bagged_tree)
```

```{r}
# Tune Bagged Tree
# Define the tuning grid
bagged_tree_grid <- grid_regular(
  tree_depth(range = c(2, 10)),
  min_n(range = c(5, 20)),
  levels = 5
)


```

Cross-Validation for Hyperparameter Tuning

```{r}
# Select Best Bagged Tree Model
#best_bagged <- bagged_tune |> select_best(metric = "rmse")

```

```{r}
# Perform cross-validation tuning
bagged_tree_tune <- bag_wkf |>
  tune_grid(
    resamples = bike_10_fold,   # Cross-validation folds
    grid = bagged_tree_grid,    # Hyperparameter grid
    metrics = metric_set(rmse, mae)  # Evaluation metrics
  )

```

Select the best model

```{r}
# Select the best hyperparameters based on RMSE
best_bagged_tree <- bagged_tree_tune |>
  select_best(metric = "rmse")

best_bagged_tree

```

Fit the model

```{r}
# Finalize the workflow with the best parameters
bagged_tree_fit <- bag_wkf |>
  finalize_workflow(best_bagged_tree) |>
  fit(data = bike_train)

```

Evaluate on Test data

```{r}
# Make predictions on the test set
bagged_tree_preds <- predict(bagged_tree_fit, new_data = bike_test) |>
  bind_cols(bike_test)

# Evaluate model performance
bagged_tree_metrics <- bagged_tree_preds |>
  metrics(truth = bike_count, estimate = .pred)

bagged_tree_metrics

```

The bagged tree model improves performance further with test RMSE around 3,050 and R² about 0.90.

```{r}

# library(vip)
# bagged_fit |> 
#   extract_fit_parsnip() |> 
#   vip::vip()

```

```{r}
# bag_final_model <- extract_fit_engine(bagged_fit)
# bag_final_model$imp |>
#  mutate(term = factor(term, levels = term)) |>
#  ggplot(aes(x = term, y = value)) +
#  geom_bar(stat ="identity") +
#  coord_flip()

# Extract the bagged tree engine fit
bag_engine <- bagged_tree_fit |>
  extract_fit_engine()

# bag_engine$imp contains the variable importance
bag_imp <- bag_engine$imp |>
  as_tibble(rownames = "term") |>
  rename(importance = 2)

ggplot(bag_imp, aes(x = reorder(term, importance), y = importance)) +
  geom_col() +
  coord_flip() +
  labs(
    x = "Variable",
    y = "Importance",
    title = "Bagged Tree Variable Importance"
  )

```

The bagged tree variable importance plot shows that a small set of predictors contributes most to the model’s predictive power.

## Random Forest

```{r}
rf_spec <- rand_forest(
  mtry = tune(), 
  trees = 500, 
  min_n = tune()
) |> 
  set_engine("ranger", importance = "impurity") |> 
  set_mode("regression")

```

Workflow

```{r}
rf_wkf <- workflow() |>
  add_recipe(MLR_rec1) |>
  add_model(rf_spec)


```

Hyperparameter Grid

```{r}
# Define the Grid for Tuning
rf_grid <- grid_regular(
  mtry(range = c(1, ncol(bike_train) - 1)), 
  min_n(range = c(2, 10)),
  levels = 5
)

```

Cross-Validation

```{r}
# Tune the Random Forest Model
rf_tune <- rf_wkf |>
  tune_grid(
    resamples = bike_10_fold,
    grid = rf_grid,
    metrics = metric_set(rmse, mae)  
  )


```

Best Hyperparameter

```{r}
best_rf <- rf_tune |>
  select_best(metric = "rmse")

# View the Best Parameters
print(best_rf)
```

workflow and fit on Training data

```{r}
# Finalize Workflow with Best Parameters
rf_fit <- rf_wkf |>
  finalize_workflow(best_rf) |>
  fit(data = bike_train)


```

Model on the Test Set

```{r}
# Make Predictions on Test Set
rf_preds <- predict(rf_fit, new_data = bike_test) |>
  bind_cols(bike_test)

# Evaluate Model Performance
rf_metrics <- rf_preds |>
  metrics(truth = bike_count, estimate = .pred) # Replace 'Rental_Count' with your target variable

# View the Metrics
print(rf_metrics)

```

The random forest gives the best test performance, with RMSE about 2,925, R² around 0.91, and MAE near 2,200 daily rentals.

##Model Comparisons on Test Set

```{r}
# LASSO model performance
metrics_set <- metric_set(rmse, mae)

lasso_preds <- predict(LASSO_final, new_data = bike_test) |>
  bind_cols(bike_test) |>
  metrics_set(truth = bike_count, estimate = .pred)
```

```{r}
# MLR model performance
# Extract the workflow from the last_fit result
final_fit_workflow <- extract_workflow(final_fit)

# Make predictions on the test set using the extracted workflow
mlr_preds <- predict(final_fit_workflow, new_data = bike_test) |>
  bind_cols(bike_test) |>
  metrics_set(truth = bike_count, estimate = .pred)

mlr_preds
```

```{r}
# Regression Tree model performance
tree_preds <- predict(tree_fit, new_data = bike_test) |>
  bind_cols(bike_test) |>
  metrics_set(truth = bike_count, estimate = .pred)
```

```{r}
# Bagged Tree model performance
bagged_tree_preds <- predict(bagged_tree_fit, new_data = bike_test) |>
  bind_cols(bike_test) |>
  metrics_set(truth = bike_count, estimate = .pred)
```

```{r}
# Random Forest model performance
rf_preds <- predict(rf_fit, new_data = bike_test) |>
  bind_cols(bike_test) |>
  metrics_set(truth = bike_count, estimate = .pred)
```

```{r}

# Combine all metrics
model_comparison <- bind_rows(
  lasso_preds |> mutate(Model = "LASSO"),
  mlr_preds |> mutate(Model = "MLR"),
  tree_preds |> mutate(Model = "Regression Tree"),
  bagged_tree_preds |> mutate(Model = "Bagged Tree"),
  rf_preds |> mutate(Model = "Random Forest")
)
model_comparison
```

```{r}
model_comparison |> 
  filter(.metric == "rmse") |> 
  select(Model, .estimate) |> 
  arrange(.estimate)

```

Comparing RMSE and MAE across models, the tree-based ensemble methods (bagged tree and random forest) clearly outperform the linear and single-tree models. 

MLR and LASSO model for final coefficient

```{r}
# MLR model final coefficients
final_fit |>
  extract_fit_parsnip() |>
  tidy()

# LASSO model final coefficients
LASSO_final |>
  extract_fit_parsnip() |>
  tidy()

```

Regression Tree Model: Plot the final fit

```{r}
# Plot the final regression tree
# Extract the fitted model from the parsnip workflow
rpart_model <- tree_fit |>
  extract_fit_parsnip()

# Plot the regression tree

par(mar = c(5, 4, 4, 2))  
rpart.plot::rpart.plot(rpart_model$fit, roundint = FALSE)


```

Bagged Tree and Random Forest Models: Produce variable importance plots

```{r}
# Bagged Tree variable importance plot
#rpart_model <- bagged_tree_fit$fit
#rpart_model$variable.importance  

```

```{r}
# Extract the bagged tree engine fit 

# Extract the bagged tree engine fit 
bag_engine <- bagged_tree_fit |>
  extract_fit_engine()

# Turn variable importance into a tibble: term, importance
bag_imp <- bag_engine$imp |>
  as_tibble(rownames = "term") |>
  rename(importance = 2)

ggplot(bag_imp, aes(x = reorder(term, importance), y = importance)) +
  geom_col() +
  coord_flip() +
  labs(
    x = "Variable",
    y = "Importance",
    title = "Bagged Tree Variable Importance"
  )


```

```{r}
# Random Forest variable importance plot

 rf_fit %>%
   extract_fit_parsnip() %>%
   vip::vip()

```

Best Model, Fit it to the Entire Dataset

```{r}
# Example: Refit the Random Forest (assuming it was the best based on the test set performance)
final_rf_fit <- rf_wkf |>
  finalize_workflow(best_rf) |>
  fit(data = bike_data)

final_rf_fit



```

Because it performed best on the test set, we refit the random forest on the full dataset to obtain our final model for predicting daily bike rent

```{r}
```

```{r}
```

```{r}
```
