---
title: "01_read_eda"
editor: visual
---

### Load Libraries

```{r}

library(readr)
library(lubridate)
library(dplyr)
library(ggplot2)

```

### Read Data

```{r}
bike_data <- read_csv(
"https://www4.stat.ncsu.edu/~online/datasets/SeoulBikeData.csv",
skip = 1,
locale = locale(encoding = "UTF-8"),
col_names = FALSE,
show_col_types = FALSE
)

colnames(bike_data) <- c(
"Date", "Rented_Bike_Count", "Hour", "Temperature",
"Humidity", "Windspeed", "Visibility", "Dew_Point_Temperature",
"Solar_Radiation", "Rainfall", "Snowfall", "Seasons",
"Holiday", "Functioning_Day"
)

head(bike_data)

```

```{r}
sum(is.na(bike_data))
str(bike_data)

```

### Clean Variables

```{r}
bike_data <- bike_data |>
mutate(
Date = dmy(Date),
Seasons = as.factor(Seasons),
Holiday = as.factor(Holiday),
Functioning_Day = as.factor(Functioning_Day)
)

str(bike_data)

```

### Summary 

```{r}
# Numeric summaries

summary(select_if(bike_data, is.numeric))

# Categorical levels

unique(bike_data$Seasons)
unique(bike_data$Holiday)
unique(bike_data$Functioning_Day)


```

### Summary by Season and Holiday

```{r}
bike_data_summary <- bike_data |>
filter(Functioning_Day == "Yes") |>
group_by(Seasons, Holiday) |>
summarize(
mean_rent   = mean(Rented_Bike_Count),
sd_rent     = sd(Rented_Bike_Count),
median_rent = median(Rented_Bike_Count),
.groups = "drop"
)

bike_data_summary


```

This table shows clear seasonal differences in bike rentals.

Summer has the highest average rentals, followed by Autumn.

Winter has very low rental counts, which matches expectations due to cold weather.

Holidays generally have slightly lower rentals than non-holidays in each season.

Spring shows more variation, with wide differences between the mean, median, and standard deviation.

Overall, season and holiday status clearly affect bike rental behavior.

### Visualization

```{r}
ggplot(bike_data_summary, aes(x = Seasons, y = mean_rent, fill = Holiday)) +
geom_bar(stat = "identity", position = "dodge") +
labs(
title = "Mean Rentals by Season and Holiday (Hour-Level)",
x = "Season",
y = "Mean Rentals"
) +
theme_minimal()


```

This plot shows clear differences in bike rentals across seasons and holidays.

Summer has the highest average rentals, followed by Autumn.

Spring rentals are moderate, and Winter rentals are the lowest.

Holidays generally have slightly lower rentals than non-holidays in every season, especially in Winter and Spring where the drop is more noticeable.

Overall, this plot shows that both season and holiday status strongly affect how many bikes are rented each hour.

### Save the dataset as an RDS file for reuse

```{r}
saveRDS(bike_data, file = "../bike_data.rds")

```
