---
title: "03_modeling_results"
editor: visual
---

### Load Libraries

```{r}
# Load libraries needed for modeling
library(tidyverse)
library(tidymodels)
library(lubridate)
```

### Load Data

```{r}
# Load dataset saved from earlier step
daily_bike <- readRDS("../daily_bike.rds")
str(daily_bike)

```

### Split Data

```{r}
set.seed(123)
bike_split <- initial_split(daily_bike, prop = 0.75, strata = Seasons)
bike_train <- training(bike_split)
bike_test  <- testing(bike_split)

```

### Recipe 1: base model 

```{r}
# This recipe prepares the data for modeling. 
# It creates a weekday and weekend variable from the date, removes the original Date column,
# standardizes all numeric predictors so they are on the same scale,
# and converts all categorical variables into dummy (0/1) variables.
# These steps help the linear model interpret the data correctly.


recipe_1 <- recipe(Rented_Bike_sum ~ ., data = bike_train) |>
step_mutate(day_type = factor(if_else(wday(Date) %in% c(1,7), "weekend", "weekday"))) |>
step_rm(Date) |>
step_normalize(all_numeric_predictors()) |>
step_dummy(all_nominal_predictors())
```

### Recipe 2: add interactions

```{r}
# This recipe builds on the first one. It includes weekday and weekend,
# removes the Date column, standardizes numeric variables,
# and makes dummy variables for categorical features.
# It also adds interaction terms so the model can capture how
# seasons and holidays work together, how seasons and temperature interact,
# and how temperature and rainfall interact. This helps the model learn
# more complex relationships in the data.


recipe_2 <- recipe(Rented_Bike_sum ~ ., data = bike_train) |>
step_mutate(day_type = factor(if_else(wday(Date) %in% c(1,7), "weekend", "weekday"))) |>
step_rm(Date) |>
step_normalize(all_numeric_predictors()) |>
step_dummy(all_nominal_predictors()) |>
step_interact(
terms = ~ starts_with("Seasons") * starts_with("Holiday") +
starts_with("Seasons") * Temperature_mean +
Temperature_mean * Rainfall_sum
)
```

### Recipe 3: add quadratic terms (polynomials) 

```{r}
# This recipe includes everything from recipe 2 (weekday and weekend,
# removing Date, standardizing numerics, dummy variables, and interactions).
# It also adds polynomial (squared) terms for several numeric predictors.
# These quadratic terms help the model capture curved relationships,
# such as rentals increasing up to a point and then leveling off.
# This makes the model more flexible for weather related effects.


recipe_3 <- recipe(Rented_Bike_sum ~ ., data = bike_train) |>
step_mutate(day_type = factor(if_else(wday(Date) %in% c(1,7), "weekend", "weekday"))) |>
step_rm(Date) |>
step_normalize(all_numeric_predictors()) |>
step_dummy(all_nominal_predictors()) |>
step_interact(
terms = ~ starts_with("Seasons") * starts_with("Holiday") +
starts_with("Seasons") * Temperature_mean +
Temperature_mean * Rainfall_sum
) |>
step_poly(
Temperature_mean,
Humidity_mean,
Windspeed_mean,
Visibility_mean,
Dew_Point_Temperature_mean,
Solar_Radiation_mean,
degree = 2
)

```

### Model Setup

```{r}
# Here I set up our linear regression model and combine it with each recipe.
# I create one workflow for each recipe so we can compare the three models:
# Model 1 uses the basic recipe, Model 2 adds interaction terms,
# and Model 3 adds both interactions and polynomial terms.



bike_mod <- linear_reg() |>
set_engine("lm")

bike_wfl_1 <- workflow() |>
add_recipe(recipe_1) |>
add_model(bike_mod)

bike_wfl_2 <- workflow() |>
add_recipe(recipe_2) |>
add_model(bike_mod)

bike_wfl_3 <- workflow() |>
add_recipe(recipe_3) |>
add_model(bike_mod)


```

### Cross-Validation

```{r}
# I use 10 fold cross validation to compare the performance of the three models.
# The training data is split into 10 parts, and each model is tested across all folds.
# Measure error using RMSE, which tells us how far the predictions are from the actual rentals.


set.seed(42)
bike_10_fold <- vfold_cv(bike_train, v = 10)

bike_res_1 <- fit_resamples(
bike_wfl_1,
resamples = bike_10_fold,
metrics = metric_set(rmse)
)

bike_res_2 <- fit_resamples(
bike_wfl_2,
resamples = bike_10_fold,
metrics = metric_set(rmse)
)

bike_res_3 <- fit_resamples(
bike_wfl_3,
resamples = bike_10_fold,
metrics = metric_set(rmse)
)


```

### Compare Metrics

```{r}

metrics_1 <- collect_metrics(bike_res_1) |> mutate(workflow = "Model 1")
metrics_2 <- collect_metrics(bike_res_2) |> mutate(workflow = "Model 2")
metrics_3 <- collect_metrics(bike_res_3) |> mutate(workflow = "Model 3")

all_metrics <- bind_rows(metrics_1, metrics_2, metrics_3)

all_metrics |> arrange(mean)

```

Model 2 has the lowest RMSE, which means it predicts daily bike rentals the most accurately.

Model 3 is slightly worse, and Model 1 performs the worst.

This shows that adding interaction terms improves the model,

but adding polynomial terms does not help much for this dataset.

### Final Fit and Results

```{r}

rmse_rows <- all_metrics |> filter(.metric == "rmse")
best_row <- rmse_rows |> arrange(mean) |> slice(1)

best_row



```

The best model chosen by cross-validation is Model 2.

Model 2 had the smallest average RMSE across the 10 folds.

This means the interaction terms in Model 2 helped the model learn important relationships

between season, holiday, temperature, and rainfall.

```{r}
workflow_list <- list(
"Model 1" = bike_wfl_1,
"Model 2" = bike_wfl_2,
"Model 3" = bike_wfl_3
)
best_wfl <- workflow_list[[ best_row$workflow ]]
final_fit <- best_wfl |>
last_fit(split = bike_split)
final_metrics <- final_fit |>
collect_metrics()
final_metrics
```

On the test data, the RMSE is about 3081 rentals, which is similar to the CV error.

This means the model generalizes well and is not overfitting.

The R-squared value is about 0.91, meaning the model explains about 91% of the variation

in daily bike rentals, which shows a strong fit.

### Coefficients

```{r}
final_model <- extract_fit_parsnip(final_fit$.workflow[[1]])
tidy(final_model)

```

Solar radiation has a large positive effect, meaning sunny days increase bike rentals.

Rainfall has a strong negative effect, and humidity and wind speed also reduce rentals.

Snowfall is not significant in this model.

Temperature becomes non-significant after including interactions, since its effect changes by season.

Spring and Winter have large negative coefficients, meaning rentals are much lower than the baseline season (Autumn).

Overall, weather and season are the main drivers of bike rental behavior.
